\pkg{DNest4}: Diffusive Nested Sampling in \proglang{C++} and \proglang{Python}

### 2018-08-07 paper/ms.tex
Paper from JSS

\documentclass[article, nojss]{jss}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{enumerate}
\usepackage{microtype}
\usepackage{dsfont}
\usepackage{pifont}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Brendon J. Brewer\\Department of Statistics\\The University of Auckland\And 
        Daniel Foreman-Mackey\\Sagan Fellow\\University of Washington}
\title{\pkg{DNest4}: Diffusive Nested Sampling in
\proglang{C++} and \proglang{Python}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Brendon J. Brewer, Daniel Foreman-Mackey} %% comma-separated
\Plaintitle{DNest4: Diffusive Nested Sampling in
C++ and Python} %% without formatting
\Shorttitle{\pkg{DNest4}: Diffusive Nested Sampling} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{In probabilistic (Bayesian) inferences, we typically want to compute
properties of the posterior distribution, describing knowledge of
unknown quantities in the context of a particular dataset and the assumed
prior information. The marginal likelihood, also known as the ``evidence'',
is a key quantity in Bayesian model selection. The Diffusive Nested Sampling algorithm, a variant of Nested Sampling, is a powerful tool for generating
posterior samples and estimating marginal likelihoods. It is effective at
solving complex problems including many where the posterior distribution is
multimodal or has strong dependencies between variables.
\pkg{DNest4} is an open source (MIT licensed),
multi-threaded implementation of this algorithm in
\proglang{C++11},
along with associated utilities including:
i)
\pkg{RJObject}, a class template for finite mixture models;
(ii)
A \proglang{Python} package allowing basic use without \proglang{C++} coding; and
iii)
Experimental support for models implemented in \proglang{Julia}.
In this paper we demonstrate \pkg{DNest4} usage through examples including
simple Bayesian data analysis, finite mixture models, and Approximate
Bayesian Computation.
}

\Keywords{bayesian inference, markov chain monte carlo,
metropolis algorithm, bayesian computation, nested sampling, \proglang{c++11},
\proglang{python}}
\Plainkeywords{bayesian inference, markov chain monte carlo,
metropolis algorithm, bayesian computation, nested sampling, c++11, python} %% without formatting
%% at least one keyword must be supplied

@@

In compact but standard notation, the posterior
posterior distribution for parameters $\params$ given data $\data$, within
the context of prior information $M$, is
\begin{align}
p(\params | \data, M) &=
\frac{p(\params | M)p(\data | \params, M)}{p(\data | M)}
\end{align}
or
\begin{align}
\textnormal{posterior} &=
\frac{\textnormal{prior} \times \textnormal{likelihood}}
     {\textnormal{marginal likelihood}}.
\end{align}

If prior information $I$ (dropped hereafter)
implies a set of possible `models' $\{M_i\}$,
rather than a single one $M$, the posterior model probabilities are given by
\begin{align}
P(M_i | \data) &=
\frac{P(M_i)p(\data | M_i)}{\sum_j P(M_j)p(\data | M_j)}
\end{align}
where
\begin{align}
p(\data | M_j) &= \int p(\theta_j | M_j)p(\data | \theta_j, M_j) \, d\theta_j
\end{align}
is the marginal likelihood of model $j$, equal to the expected value of the
likelihood function with respect to the prior distribution.
This kind of calculation is often
called ``model selection'' or ``model averaging'', and the results
are often presented as ratios
of marginal likelihoods, known as Bayes Factors.
When discussing computational matters, the prior distribution for parameters
is
often written $\pi(\theta)$, the likelihood $L(\theta)$,
and the marginal likelihood $Z$. A popular alternative name for the marginal
likelihood, which emphasizes its role in Bayesian model averaging,
is the ``evidence''.

Nested Sampling \citep[NS;][]{skilling2006nested} is a Monte Carlo method whose main
aim is to calculate $Z$. However, it can also be used to generate samples
to represent the posterior distribution $\pi(\theta)L(\theta)/Z$, or
any other distribution proportional to
$\pi(\theta)\Phi\left[L(\theta)\right]$ where $\Phi$ is any monotonic function.
This latter property makes NS particularly useful for statistical mechanics
calculations \citep{partay2010efficient, baldock2016determining},
where the
``canonical'' family of distributions proportional to
$\pi(\theta)L(\theta)^\beta$ is of interest. In such applications,
$L(\theta)$ is usually equivalent to
$\exp(-\textnormal{energy})$. NS is particularly efficient for this, since
only a single run (exploring regions of higher and higher $L$) is needed, and different canonical
distributions can be obtained by re-weighting the output points.

A defining feature of NS is that it works with a sequence of
{\em constrained prior} distributions, proportional to $\pi$ but
restricted to regions of the parameter space where $L(\theta)$
is above some threshold $\ell$:
\begin{align}
p(\theta; \ell) &=
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell\right]}{X(\ell)}
\label{eqn:constrained_prior}
\end{align}
where
\begin{align}
X(\ell) &= \int \pi(\theta) \mathds{1}\left[L(\theta) > \ell\right] \, d\theta
\end{align}
is the amount of prior mass which has likelihood greater than $\ell$, and
$\mathds{1}()$ is the indicator function which takes the value 1 if the
argument is true and 0 otherwise.
In the standard NS framework, the sequence of $\ell$ values is selected
so that $X(\ell)$ shrinks by a factor $\approx e^{-1/N}$ per iteration, where
$N$ is the number of particles used. This geometric compression of the
parameter space is a defining feature of NS.

Sampling from the constrained priors (Equation~\ref{eqn:constrained_prior})
is required, and Markov Chain Monte Carlo (MCMC) is a popular method of doing
this, although alternatives exist
\citep[e.g.,][]{feroz2009multinest, handley2015polychord}.
Diffusive Nested Sampling \citep[DNS;][]{brewer2011diffusive}
is an alternative to NS for
problems where MCMC is the only viable sampling method. DNS is based on the
Metropolis algorithm, and evolves one or more particles in the parameter space, along with
an integer index variable $j$ for each particle,
to explore the following joint distribution:
\begin{align}
p(\theta, j) &= p(j)p(\theta | j)\\
&= w_j \times
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
\label{eqn:target_distribution}
\end{align}
where the $\{\ell_j\}$ are a sequence of increasing likelihood thresholds
or {\em levels},
$\ell_0 = 0$, and $\{w_j\}$ is the marginal distribution for $j$.
The marginal distribution for $\theta$ is then a {\em mixture} of
constrained priors:
\begin{align}
p(\theta) &=
\pi(\theta)\sum_{j=0}^{j_{\rm max}}
\frac{w_j\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
\label{eqn:mixture_of_constrained_priors}
\end{align}
The DNS algorithm consists of two stages. In the first stage,
the particle(s) are initialized from the prior $\pi(\theta)$, equivalent
to the mixture of constrained priors with a single level whose log likelihood
threshold is $-\infty$. The mixture of constrained priors evolves by adding new levels,
each of which compresses the distribution by a factor of about $e \approx 2.71818$
(see \citet{brewer2011diffusive} for details). In this stage,
the mixture weights $\{w_j\}$ are set according to
\begin{align}
w_j &\propto \exp(j/\lambda),\label{eqn:weighting}
\end{align}
where $\lambda$ is a scale length.
This enhances the probability that particles stay close to the
highest likelihood regions seen.
The second stage sets the mixture weights to be approximately
uniform ($w_j \propto 1$), with some tweaks described by
\citet{brewer2011diffusive}.

The mixture of constrained priors tends to be easier to sample than the
posterior, as the prior is always a mixture component, allowing the
MCMC chain to mix between different modes in some circumstances. The marginal
likelihood estimate can also be more accurate than standard MCMC-based NS
\citep{brewer2011diffusive}, as less information is discarded.
DNS has been applied several times in astrophysics
\citep[e.g.,][]{pancoast2014modelling, huppenkothen2015dissecting,
brewer2015fast}
and was recently used in a biological application
\citep{dybowski2015single}.

@@

Alternatively, the \proglang{Python}
bindings (Section~\ref{sec:python_bindings}) allow you
to specify a model class in \proglang{Python}, and run \pkg{DNest4} entirely
in the \proglang{Python} interpreter without having to invoke the \proglang{C++}
compiler.

\section{Relation to other algorithms}\label{sec:relation}
The Diffusive Nested Sampling algorithm, and its implementation in
\pkg{DNest4}, has advantages and disadvantages compared to other Bayesian
computation algorithms and software.

In the Nested Sampling sphere,
the simple \proglang{C} implementation of given by
\citet{skilling2006nested} is useful for understanding the classic NS
algorithm. Similarly to \pkg{DNest4}, it is the user's responsibility
to implement the MCMC moves used. The DNS algorithm produces more
accurate estimates of $Z$ and ought to outperform classic NS on multimodal
problems.

\pkg{MultiNest} \citep{feroz2009multinest} is a
\proglang{Fortran}
Nested Sampling implementation
that dispenses with MCMC for generating new particles from constrained
prior distributions. Instead, the new particles are generated by making
an approximation to the constrained prior using ellipsoids. This tends to work
well in low to moderate dimensional parameter spaces (up to $\sim 40$).
In higher dimensions, \pkg{DNest4} is more useful than \pkg{MultiNest}
since it is based on MCMC. We expect \pkg{MultiNest} to be more useful
than
\pkg{DNest4} only on low to moderate dimensional problems with slow
likelihood evaluations, where \pkg{DNest4}'s reliance on MCMC becomes costly.
The more recent \pkg{POLYCHORD} \citep{handley2015polychord} combines
\pkg{MultiNest}-like methods with MCMC.

Outside of Nested Sampling,
the popular \pkg{JAGS} \citep{jags} and \pkg{Stan} \citep{stan} packages
are more convenient than \pkg{DNest4} for specifying models, by virtue
of their model-specification languages. For users who are interested only
in the posterior distribution and do not need the marginal likelihood,
these are very useful.
However, being based on Gibbs
sampling and Hamiltonian MCMC respectively, they can run into difficulty
on multimodal posterior distributions.

\pkg{emcee} \citep{emcee} is a popular \proglang{Python} package for MCMC
based on affine-invariant ensemble sampling. The user needs only to
specify a \proglang{Python} function evaluating the log of the posterior
density (up to a normalizing constant). This user-friendliness is a key
advantage of \pkg{emcee}, and its algorithm performs well on low--moderate
dimensional (up to $\sim 50$) parameter spaces with a unimodel
target distribution (which can be highly dependent). However, it
can give misleading results in higher dimensions \citep{huijser}.

Algorithms based on ``annealing'' or ``tempering'', such as parallel
tempering \citep{hansmann1997parallel} and annealed importance sampling
\citep{neal2001annealed} are
related to Nested Sampling and are useful on multimodal posterior distributions.
However, they require much more tuning than NS (in the form of an
{\em annealing schedule}) and do
not work on phase change problems, unlike Nested Sampling
\citep{skilling2006nested}.

@@

%\begin{table}
%
%  \begin{tabular}{cc}
%    Knuth & Lamport
%  \end{tabular}}
%\end{table}

\begin{table}
\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
Package     &   Easy to     & High   &   Multimodal     & Dependent   &   Phase   &   Computes $Z$?\\
            &   implement & dimensions? & distributions? & distributions? & changes? & \\
& models? & & & & &\\
\hline
\pkg{DNest4}    & \xmark  & \cmark & \cmark & \cmark &\cmark &\cmark\\
\pkg{emcee}    & \cmark  & \xmark & \xmark & \cmark &\xmark &\xmark\\
\pkg{JAGS}    & \cmark  & \cmark & \xmark & \xmark & \xmark &\xmark\\
\pkg{MultiNest}    & \cmark  & \xmark & \cmark & \cmark & \cmark &\cmark\\
\pkg{Stan}    & \cmark  & \cmark & \xmark & \cmark & \xmark &\xmark\\
\hline
\end{tabular}}
\caption{A simplified summary of the advantages and disadvantages of some
Bayesian computation software packages.\label{tab:packages}}
\end{center}
\end{table}

\section{Markov chain monte carlo}\label{sec:mcmc}
DNS is build upon the Metropolis-Hastings algorithm.
In this algorithm, the acceptance probability $\alpha$
is given by
\begin{align}
\alpha &= \min\left(1,
\frac{q(\params'|\params)}{q(\params | \params')}
\frac{\pi(\params')}{\pi(\params)}\frac{L(\params')}{L(\params)}
\right)
\end{align}
where $q(\theta' | \theta)$ is the proposal distribution used to generate
a new position $\theta'$ from the current position $\theta$. Often,
$q$ is {\em symmetric} so the $q$ terms cancel in the acceptance probability.

In DNS, the target distribution is not the posterior but rather
the joint distribution in Equation~\ref{eqn:target_distribution}.
Moves of $\theta$ are done keeping $j$ fixed, so we only need
to consider the Metropolis acceptance probability for fixed $j$,
i.e., with respect to a single constrained prior like
Equation~\ref{eqn:constrained_prior}.
Hence, the appropriate acceptance probability
for a proposed move from $\theta$ to $\theta'$ is
\begin{align}
\alpha &= \min\left[1,
\frac{q(\params'|\params)}{q(\params | \params')}
\frac{\pi(\params')}{\pi(\params)}
\mathds{1}\left(L(\params') > \ell_j\right)
\right]
\label{eqn:log_hastings}
\end{align}
where $\ell_j$ is the likelihood threshold for the current level $j$.
There are also moves that propose a change to $j$ while keeping
$\params$ fixed, but the details are less relevant to the user.

For convenience later on, we 
separate the prior and proposal-related terms from
the likelihood-related term, and write the former as
\begin{align}
H = \frac{q(\params'|\params)}{q(\params | \params')}
\times \frac{\pi(\params')}{\pi(\params)}.
\end{align}

The logarithm of $H$,
\begin{align}
\ln(H) = \ln\left[\frac{q(\params'|\params)}{q(\params | \params')}
\times \frac{\pi(\params')}{\pi(\params)}\right],
\end{align}
is the user's responsibility when implementing models,
and will become relevant in Sections~\ref{sec:perturb}
and~\ref{sec:proposals}.
In terms of $\ln(H)$, the
acceptance probability becomes
\begin{align}
\alpha &= \min\left[1,
e^{\ln(H)}\times
\mathds{1}\left(L(\params') > \ell_j\right)
\right].\label{eqn:logH}
\end{align}

%The \pkg{DNest4} \code{Sampler} class handles the likelihood check,
%while the \code{perturb(DNest4::RNG\&)} member function handles the
%rest of the terms. The value returned from \code{perturb(DNest4::RNG\&)}
%must be
%for your problem.

%\subsection{Diffusive Nested Sampling}

@@

Development of \pkg{DNest4} takes place in the \pkg{git} repository located at
\begin{center}
\url{https://github.com/eggplantbren/DNest4/}
\end{center}
The software
is licensed under the permissive open source
MIT license. To compile and run \pkg{DNest4},
you require a recent version of the GNU
\proglang{C++} compiler, \pkg{g++} \citep{gcc}.
\pkg{DNest4} uses features from the \proglang{C++11} standard
\citep{c++11}.

The \proglang{Python}
packages \pkg{NumPy} \citep{numpy}, \pkg{matplotlib} \citep{matplotlib},
and \pkg{Cython} \citep{cython} are also needed.
To download and compile \pkg{DNest4}
the following steps are sufficient:
\begin{CodeChunk}
\begin{CodeInput}
> wget https://github.com/eggplantbren/DNest4/archive/0.1.5.tar.gz
> tar xvzf 0.1.5.tar.gz
> mv DNest4-0.1.5 DNest4
> cd DNest4/code
> make
> cd ../python
> python setup.py install
\end{CodeInput}
\end{CodeChunk}

In Mac OS X, the final line (which installs the \proglang{Python} parts)
of \pkg{DNest4})
needs to provide information about
your OS version. For example, if your computer runs Mac OS X 10.9,
the installation command for the \proglang{Python} package is
\begin{CodeChunk}
\begin{CodeInput}
> MACOSX_DEPLOYMENT_TARGET=10.9 python setup.py install
\end{CodeInput}
\end{CodeChunk}

We recommend you create an environment variable called \code{DNEST4\_PATH}
and set it to the directory {\em above} the \pkg{DNest4} directory. Then,
if the model templates from the \pkg{DNest4} repository are copied
to any other location on your system and used as the basis for
new work, their Makefiles will continue to function.

\section[Running DNest4]{Running \pkg{DNest4}}\label{sec:running}
To demonstrate \pkg{DNest4},
we will use a simple linear regression example where the
sampling distribution is
\begin{align}
y_i | m, b, \sigma &\sim \textnormal{Normal}(mx_i + b, \sigma^2)
\end{align}
and the priors are
\begin{align}
m &\sim \textnormal{Normal}(0, 1000^2)\\
b &\sim \textnormal{Normal}(0, 1000^2)\\
\ln\sigma &\sim \textnormal{Uniform}(-10, 10).
\end{align}
These are naive diffuse priors and do not have any special status.
The dataset is shown in Figure~\ref{fig:regression_lines}
and the code is included in the {\tt code/Examples/StraightLine}
subdirectory, and a slightly simplified version of the
code is explained in Section~\ref{sec:models}.
To execute \pkg{DNest4} on this problem, go to this directory and
execute {\tt main}. The output to the screen
should contain information about levels and ``saving particles to disk''.
After 10,000 particles have been saved to the disk, the run will terminate.

\section{Output files}
The executable {\tt main} is responsible for the exploration part of the
DNS algorithm (i.e., running the MCMC chain, building
levels, and then exploring all the levels). It creates three text output files,
{\tt sample.txt}, {\tt sample\_info.txt}, and {\tt levels.txt}.

>>>>

\documentclass[article]{jss}
\usepackage{thumbpdf,lmodern} 
\graphicspath{{Figures/}}

\usepackage{amsmath}
\usepackage{enumerate}
\usepackage{dsfont}
\usepackage{pifont}

@@

\author{Brendon J. Brewer\\Department of Statistics\\The University of Auckland\And 
        Daniel Foreman-Mackey\\Sagan Fellow\\University of Washington}
\title{\pkg{DNest4}: Diffusive Nested Sampling in
\proglang{C++} and \proglang{Python}}

\Plainauthor{Brendon J. Brewer, Daniel Foreman-Mackey}

\Plaintitle{DNest4: Diffusive Nested Sampling in C++ and Python}

\Shorttitle{\pkg{DNest4}: Diffusive Nested Sampling}

\Abstract{In probabilistic (Bayesian) inferences, we typically want to
  compute properties of the posterior distribution, describing
  knowledge of unknown quantities in the context of a particular
  dataset and the assumed prior information. The marginal likelihood,
  also known as the ``evidence'', is a key quantity in Bayesian model
  selection. The diffusive nested sampling algorithm, a variant of
  nested sampling, is a powerful tool for generating posterior samples
  and estimating marginal likelihoods. It is effective at solving
  complex problems including many where the posterior distribution is
  multimodal or has strong dependencies between variables.
  \pkg{DNest4} is an open source (MIT licensed), multi-threaded
  implementation of this algorithm in \proglang{C++11}, along with
  associated utilities including: i) `\code{RJObject}', a class template
  for finite mixture models; (ii) a \proglang{Python} package allowing
  basic use without \proglang{C++} coding; and iii) experimental
  support for models implemented in \proglang{Julia}.  In this paper
  we demonstrate \pkg{DNest4} usage through examples including simple
  Bayesian data analysis, finite mixture models, and approximate
  Bayesian computation.  }

\Keywords{Bayesian inference, Markov chain Monte Carlo, Metropolis
  algorithm, Bayesian computation, nested sampling, \proglang{C++11},
  \proglang{Python}}

\Plainkeywords{Bayesian inference, Markov chain Monte Carlo,
Metropolis algorithm, bayesian computation, nested sampling, C++11, Python} %% without formatting
%% at least one keyword must be supplied

@@

In compact but standard notation, the posterior distribution for
parameters $\params$ given data $\data$, within the context of prior
information $M$, is
\begin{align*}
p(\params | \data, M) &=
\frac{p(\params | M)p(\data | \params, M)}{p(\data | M)}
\end{align*}
or
\begin{align*}
\textnormal{posterior} &=
\frac{\textnormal{prior} \times \textnormal{likelihood}}
     {\textnormal{marginal likelihood}}.
\end{align*}

If prior information $I$ (dropped hereafter)
implies a set of possible ``models'' $\{M_i\}$,
rather than a single one $M$, the posterior model probabilities are given by
\begin{align*}
P(M_i | \data) &=
\frac{P(M_i)p(\data | M_i)}{\sum_j P(M_j)p(\data | M_j)},
\end{align*}
where
\begin{align*}
p(\data | M_j) &= \int p(\theta_j | M_j)p(\data | \theta_j, M_j) \, d\theta_j
\end{align*}
is the marginal likelihood of model $j$, equal to the expected value of the
likelihood function with respect to the prior distribution.
This kind of calculation is often
called ``model selection'' or ``model averaging'', and the results
are often presented as ratios
of marginal likelihoods, known as Bayes factors.
When discussing computational matters, the prior distribution for parameters
is
often written $\pi(\theta)$, the likelihood $L(\theta)$,
and the marginal likelihood $Z$. A popular alternative name for the marginal
likelihood, which emphasizes its role in Bayesian model averaging,
is the ``evidence''.

Nested sampling \citep[NS;][]{skilling2006nested} is a Monte Carlo
method whose main aim is to calculate $Z$. However, it can also be
used to generate samples to represent the posterior distribution
$\pi(\theta)L(\theta)/Z$, or any other distribution proportional to
$\pi(\theta)\Phi\left[L(\theta)\right]$ where $\Phi$ is any monotonic
function.  This latter property makes nested sampling particularly useful for
statistical mechanics calculations \citep{partay2010efficient,
  baldock2016determining}, where the ``canonical'' family of
distributions proportional to $\pi(\theta)L(\theta)^\beta$ is of
interest. In such applications, $L(\theta)$ is usually equivalent to
$\exp(-\textnormal{energy})$. Nested sampling is particularly
efficient for this, since only a single run (exploring regions of
higher and higher $L$) is needed, and different canonical
distributions can be obtained by re-weighting the output points.

A defining feature of nested sampling is that it works with a sequence of
{\em constrained prior} distributions, proportional to $\pi$ but
restricted to regions of the parameter space where $L(\theta)$
is above some threshold $\ell$:
\begin{align}
p(\theta; \ell) &=
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell\right]}{X(\ell)},
\label{eqn:constrained_prior}
\end{align}
where
\begin{align*}
X(\ell) &= \int \pi(\theta) \mathds{1}\left[L(\theta) > \ell\right] \, d\theta
\end{align*}
is the amount of prior mass which has likelihood greater than $\ell$, and
$\mathds{1}()$ is the indicator function which takes the value 1 if the
argument is true and 0 otherwise.
In the standard nested sampling framework, the sequence of $\ell$ values is selected
so that $X(\ell)$ shrinks by a factor $\approx e^{-1/N}$ per iteration, where
$N$ is the number of particles used. This geometric compression of the
parameter space is a defining feature of nested sampling.

Sampling from the constrained priors (Equation~\ref{eqn:constrained_prior})
is required, and Markov chain Monte Carlo (MCMC) is a popular method of doing
this, although alternatives exist
\citep[e.g.,][]{feroz2009multinest, handley2015polychord}.
Diffusive nested sampling \citep[DNS;][]{brewer2011diffusive}
is an alternative to nested sampling for
problems where MCMC is the only viable sampling method. DNS is based on the
Metropolis algorithm, and evolves one or more particles in the parameter space, along with
an integer index variable $j$ for each particle,
to explore the following joint distribution:
\begin{align}
p(\theta, j) &= p(j)p(\theta | j)\nonumber\\
&= w_j \times
\frac{\pi(\theta)\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)},
\label{eqn:target_distribution}
\end{align}
where the $\{\ell_j\}$ are a sequence of increasing likelihood thresholds
or {\em levels},
$\ell_0 = 0$, and $\{w_j\}$ is the marginal distribution for $j$.
The marginal distribution for $\theta$ is then a {\em mixture} of
constrained priors:
\begin{align}
p(\theta) &=
\pi(\theta)\sum_{j=0}^{j_{\rm max}}
\frac{w_j\mathds{1}\left[L(\theta) > \ell_j\right]}{X(\ell_j)}.
\label{eqn:mixture_of_constrained_priors}
\end{align}
The DNS algorithm consists of two stages. In the first stage,
the particle(s) are initialized from the prior $\pi(\theta)$, equivalent
to the mixture of constrained priors with a single level whose log-likelihood
threshold is $-\infty$. The mixture of constrained priors evolves by adding new levels,
each of which compresses the distribution by a factor of about $e \approx 2.71818$
(see \citet{brewer2011diffusive} for details). In this stage,
the mixture weights $\{w_j\}$ are set according to
\begin{align}
w_j &\propto \exp(j/\lambda),\label{eqn:weighting}
\end{align}
where $\lambda$ is a scale length.
This enhances the probability that particles stay close to the
highest likelihood regions seen.
The second stage sets the mixture weights to be approximately
uniform ($w_j \propto 1$), with some tweaks described by
\citet{brewer2011diffusive}.

The mixture of constrained priors tends to be easier to sample than the
posterior, as the prior is always a mixture component, allowing the
MCMC chain to mix between different modes in some circumstances. The marginal
likelihood estimate can also be more accurate than standard MCMC-based nested sampling
\citep{brewer2011diffusive}, as less information is discarded.
DNS has been applied several times in astrophysics
\citep[e.g.,][]{pancoast2014modelling, huppenkothen2015dissecting,
brewer2015fast}
and was recently used in a biological application
\citep{dybowski2015single}.

@@

Alternatively, the \proglang{Python} \citep{python}
bindings (Section~\ref{sec:python_bindings}) allow you
to specify a model class in \proglang{Python}, and run \pkg{DNest4} entirely
in the \proglang{Python} interpreter without having to invoke the \proglang{C++}
compiler.

\section{Relation to other algorithms}\label{sec:relation}
The diffusive nested sampling algorithm, and its implementation in
\pkg{DNest4}, has advantages and disadvantages compared to other Bayesian
computation algorithms and software.

In the nested sampling sphere,
the simple \proglang{C} implementation of given by
\citet{skilling2006nested} is useful for understanding the classic nested sampling
algorithm. Similarly to \pkg{DNest4}, it is the user's responsibility
to implement the MCMC moves used. The DNS algorithm produces more
accurate estimates of $Z$ and ought to outperform classic nested sampling on multimodal
problems.

\pkg{MultiNest} \citep{feroz2009multinest} is a
\proglang{Fortran}
nested sampling implementation
that dispenses with MCMC for generating new particles from constrained
prior distributions. Instead, the new particles are generated by making
an approximation to the constrained prior using ellipsoids. This tends to work
well in low to moderate dimensional parameter spaces (up to $\sim 40$).
In higher dimensions, \pkg{DNest4} is more useful than \pkg{MultiNest}
since it is based on MCMC. We expect \pkg{MultiNest} to be more useful
than
\pkg{DNest4} only on low to moderate dimensional problems with slow
likelihood evaluations, where \pkg{DNest4}'s reliance on MCMC becomes costly.
The more recent \pkg{POLYCHORD} \citep{handley2015polychord} combines
\pkg{MultiNest}-like methods with MCMC.

Outside of nested sampling,
the popular \pkg{JAGS} \citep{jags} and \pkg{Stan} \citep{stan} packages
are more convenient than \pkg{DNest4} for specifying models, by virtue
of their model-specification languages. For users who are interested only
in the posterior distribution and do not need the marginal likelihood,
these are very useful.
However, being based on Gibbs
sampling and Hamiltonian MCMC respectively, they can run into difficulty
on multimodal posterior distributions.

\pkg{emcee} \citep{emcee} is a popular \proglang{Python} package for MCMC
based on affine-invariant ensemble sampling. The user needs only to
specify a \proglang{Python} function evaluating the log of the posterior
density (up to a normalizing constant). This user-friendliness is a key
advantage of \pkg{emcee}, and its algorithm performs well on low to moderate
dimensional (up to $\sim 50$) parameter spaces with a unimodel
target distribution (which can be highly dependent). However, it
can give misleading results in higher dimensions \citep{huijser}.

Algorithms based on ``annealing'' or ``tempering'', such as parallel
tempering \citep{hansmann1997parallel} and annealed importance sampling
\citep{neal2001annealed} are
related to nested sampling and are useful on multimodal posterior distributions.
However, they require much more tuning than nested sampling (in the form of an
{\em annealing schedule}) and do
not work on phase change problems, unlike nested sampling
\citep{skilling2006nested}.

@@

\begin{table}[t!]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
\hline
Package     &   Easy to     & High   &   Multimodal     & Dependent   &   Phase   &   Computes $Z$?\\
            &   implement & dimensions? & distributions? & distributions? & changes? & \\
& models? & & & & &\\
\hline
\pkg{DNest4}    & \xmark  & \cmark & \cmark & \cmark &\cmark &\cmark\\
\pkg{emcee}    & \cmark  & \xmark & \xmark & \cmark &\xmark &\xmark\\
\pkg{JAGS}    & \cmark  & \cmark & \xmark & \xmark & \xmark &\xmark\\
\pkg{MultiNest}    & \cmark  & \xmark & \cmark & \cmark & \cmark &\cmark\\
\pkg{Stan}    & \cmark  & \cmark & \xmark & \cmark & \xmark &\xmark\\
\hline
\end{tabular}}
\caption{A simplified summary of the advantages and disadvantages of some
Bayesian computation software packages.\label{tab:packages}}
\end{table}
\newpage
\section{Markov chain Monte Carlo}\label{sec:mcmc}
DNS is build upon the Metropolis-Hastings algorithm.
In this algorithm, the acceptance probability $\alpha$
is given by
\begin{align*}
\alpha &= \min\left(1,
\frac{q(\params'|\params)}{q(\params | \params')}
\frac{\pi(\params')}{\pi(\params)}\frac{L(\params')}{L(\params)}
\right),
\end{align*}
where $q(\theta' | \theta)$ is the proposal distribution used to generate
a new position $\theta'$ from the current position $\theta$. Often,
$q$ is {\em symmetric} so that the $q$ terms cancel in the acceptance probability.

In DNS, the target distribution is not the posterior but rather
the joint distribution in Equation~\ref{eqn:target_distribution}.
Moves of $\theta$ are done keeping $j$ fixed, so we only need
to consider the Metropolis acceptance probability for fixed $j$,
i.e., with respect to a single constrained prior like
Equation~\ref{eqn:constrained_prior}.
Hence, the appropriate acceptance probability
for a proposed move from $\theta$ to $\theta'$ is
\begin{align}
\alpha &= \min\left[1,
\frac{q(\params'|\params)}{q(\params | \params')}
\frac{\pi(\params')}{\pi(\params)}
\mathds{1}\left(L(\params') > \ell_j\right)
\right],
\label{eqn:log_hastings}
\end{align}
where $\ell_j$ is the likelihood threshold for the current level $j$.
There are also moves that propose a change to $j$ while keeping
$\params$ fixed, but the details are less relevant to the user.

For convenience later on, we 
separate the prior and proposal-related terms from
the likelihood-related term, and write the former as
\begin{align*}
H = \frac{q(\params'|\params)}{q(\params | \params')}
\times \frac{\pi(\params')}{\pi(\params)}.
\end{align*}

The logarithm of $H$,
\begin{align*}
\ln(H) = \ln\left[\frac{q(\params'|\params)}{q(\params | \params')}
\times \frac{\pi(\params')}{\pi(\params)}\right],
\end{align*}
is the user's responsibility when implementing models,
and will become relevant in Sections~\ref{sec:perturb}
and~\ref{sec:proposals}.
In terms of $\ln(H)$, the
acceptance probability becomes
\begin{align}
\alpha &= \min\left[1,
e^{\ln(H)}\times
\mathds{1}\left(L(\params') > \ell_j\right)
\right].\label{eqn:logH}
\end{align}

@@

Development of \pkg{DNest4} takes place in the git repository located at
\url{https://github.com/eggplantbren/DNest4/}.
The software
is licensed under the permissive open source
MIT license. To compile and run \pkg{DNest4},
you require a recent version of the GNU
\proglang{C++} compiler, \pkg{g++} \citep{gcc}.
\pkg{DNest4} uses features from the \proglang{C++11} standard
\citep{c++11}.

The \proglang{Python}
packages \pkg{NumPy} \citep{numpy}, \pkg{matplotlib} \citep{matplotlib},
and \pkg{Cython} \citep{cython} are also needed.
To download and compile \pkg{DNest4}
the following steps are sufficient:
%
\begin{CodeChunk}
\begin{CodeInput}
$ wget https://github.com/eggplantbren/DNest4/archive/0.1.4.tar.gz
$ tar xvzf 0.1.4.tar.gz
$ mv DNest4-0.1.4 DNest4
$ cd DNest4/code
$ make
$ cd ../python
$ python setup.py install
\end{CodeInput}
\end{CodeChunk}
%
In Mac OS X, the final line (which installs the \proglang{Python} parts
of \pkg{DNest4})
needs to provide information about
your OS version. For example, if your computer runs Mac OS X 10.9,
the installation command for the \proglang{Python} package is:
%
\begin{CodeChunk}
\begin{CodeInput}
$ MACOSX_DEPLOYMENT_TARGET=10.9 python setup.py install
\end{CodeInput}
\end{CodeChunk}
%
We recommend to create an environment variable called \code{DNEST4\_PATH}
and set it to the directory {\em above} the \pkg{DNest4} directory. Then,
if the model templates from the \pkg{DNest4} repository are copied
to any other location on your system and used as the basis for
new work, their Makefiles will continue to function.

\section[Running DNest4]{Running \pkg{DNest4}}\label{sec:running}
To demonstrate \pkg{DNest4},
we will use a simple linear regression example where the
sampling distribution is
\begin{align*}
y_i | m, b, \sigma &\sim \textnormal{Normal}(mx_i + b, \sigma^2)
\end{align*}
and the priors are
\begin{align*}
m &\sim \textnormal{Normal}(0, 1000^2),\\
b &\sim \textnormal{Normal}(0, 1000^2),\\
\ln\sigma &\sim \textnormal{Uniform}(-10, 10).
\end{align*}
These are na{\"i}ve diffuse priors and do not have any special status.
The dataset is shown in Figure~\ref{fig:fig3} on the right
and the code is included in the \code{code/Examples/StraightLine}
subdirectory, and a slightly simplified version of the
code is explained in Section~\ref{sec:models}.
To execute \pkg{DNest4} on this problem, go to this directory and
execute \code{main}. The output to the screen
should contain information about levels and ``saving particles to disk''.
After 10,000 particles have been saved to the disk, the run will terminate.

\section{Output files}
The executable \code{main} is responsible for the exploration part of the
DNS algorithm (i.e., running the MCMC chain, building
levels, and then exploring all the levels). It creates three text output files,
\code{sample.txt}, \code{sample\_info.txt}, and \code{levels.txt}.

The first output
file, \code{sample.txt}, contains a sampling of parameter values that
represents the {\it mixture of constrained priors} (the target distribution
used in DNS), \emph{not} the
posterior distribution. Each line of \code{sample.txt} represents a point in
parameter space. In the linear example, there are three parameters
($m$, $b$, and $\sigma$); so there
are three columns in \code{sample.txt}.
Each time a point is saved to \code{sample.txt}, \pkg{DNest4} prints
the message ``Saving a particle to disk. N = ...''.
